A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}

### 2023-11-07 paper.tex
txs auto checkin

This document provides a conceptual solution for a specific aspect: heritage trail signage. The solution is conceptual in that it tries to follow a guideline that takes into account many aspects and provides a practical way that is adaptable to other environments.

{\huge Anticipare qui uso dei QR code}

@@

The functions of signage are two:
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map. The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream". A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}). The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
\item a sign should explain the reasons for interest in a site. This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}). In this sense, the CAI signal above is not sufficient. A simple board with a site name fits only when the site is sufficiently popular. Otherwise, a more structured message should be used to explain its relevance.
\end{itemize}

@@

@@

In Italy, a municipality is the smallest geographical-administrative unit offering accessible data about tourism, which are the fundamentals of our research. The municipality of \emph{Bagni di Lucca} boasts historical popularity for its landscapes told by famous poets (such as Giosuè Carducci, Giovanni Pascoli, Eugenio Montale, George Gordon Byron, Percy Bysshe Shelley) and celebrities like Paolina Bonaparte. The typical Liberty style architecture characterizes the villas like the \textit{Real Casinò} featuring elegant gardens representing the power of the Republic of \emph{Lucca} in the early XIX century and the \emph{Belle Epoque}. Such historical prominence made the town a well-known destination abroad, thanks to the massive emigration and the presence of the Thermal Bath structure \footnote{Terme di Bagni di Lucca: \url{https://www.termebagnidilucca.it/}}, nearby must-see attractions of the time.

@@

\begin{table}
    \footnotesize
    \setstretch{1,2}
    \centering
    \begin{tabular}{l||l|l} \hline 
        {\bf Region} & {\bf Bagni di Lucca} & {\bf Lucca} \\ \hline
        Resident density (residents/{$km^2$}) & 33.95 & 215.71 \\
        Touristic density (visitors/{$km^2$}) & 106.41 & 1866.96 \\
        Average stay  (days) &  2.98 &  3.36 \\
        Accommodation density (accommodations/{$km^2$}) & 10.14 & 19.30 \\
        Coexistence Index (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
         \hline
    \end{tabular}
    \caption{Tourist numbers in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
    \label{tab:tourism}
\end{table}

@@

Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade. Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley. Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long yet incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area. The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.

From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).

After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.

Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area.

%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.

\begin{table}
    \centering
    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
         {\bf Touristic promotion }&  {\bf Kind of service}\\
         \hline \hline 
         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
         \hline 
         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
         \hline 
         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
        \hline 
         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
         \hline 
         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
         \hline 
         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
         \hline 
        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
        \hline
    \end{tabular}
    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players. The companies and associations in the list have been individually contacted but some did not respond to our survey on touristic information.}}}
    \label{tab:experience}
\end{table}

The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality. It is 7,22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters of altitude.

\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}. Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts. Definitely, despite its variety of tourist attractions, \emph{Casoli} exhibits unique features in its comprehensive tourism.

%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.

Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways. According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities. Local stakeholders operating in the tourism system offer a variety of services all year round. Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability. As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination. Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric. The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.

@@

As explained in the previous section, the area around \textit{Casoli} is a suitable case study location due to its long history, which is reflected in its architectural and geological features. In the following, we define the relevant features and report on their accessibility.

The cave named {\em Buca La Piella} is the foremost site from an archeological point of view. It is located in a fascinating environment outside conventional circuits, which is why it is less frequented and consequently less compromised. It is reachable leaving a marked path to follow the bed of a tributary of the \emph{Lima} stream. The trail is dotted with minor karst caves, and the {\em Buca La Piella} is reachable by climbing up on a steep slope in the wood on the left of the watercourse.

@@

    \item URL LoR: such information is useful only in a few locations since the area is not uniformly covered by broadband networks: for instance, during a recon to the {\em Piella} cave, none of the smartphones of the participants received sufficient broadband network signal to visit the linked page. However, the application records the URL so that the user can visit it when entering a covered area. Each Web page contains site-specific information, an interactive map with the location of all the QR tags (see figure \ref{fig:linkmap}), and a form for user feedback (see figure \ref{fig:linkpoll}). The map is hosted by the UMap Web service (\url{https://umap.openstreetmap.fr/it/}) \citeWeb{umap}, and displays the location and content of all QR-tags (as shown in figure \ref{fig:qrcodecontent}). The server hosting the site is configured to collect aggregated statistics about visitors' activity.
\end{itemize}

@@

Its {\em performance} is currently difficult to evaluate. The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons, one being that the QR tag cannot record or transmit read events. Another is that the user may be unable or unwilling to visit the URL in the tag. Finally, the URL can be visited by means different from the QR reading. For such reasons, despite we acknowledging its importance for the stakeholders, we do not discuss such an index.

@@

@@

A different application-based solution privileging passive devices is based on a writable NFC tag close to the one delivering the content. To update the number of visitors that have read the informational NFC, the user application reads and increments the count recorded on the writable NFC. Here too the primary drawback compared with our QR-tag based solution is the need for an ad-hoc application. In addition the event of data pollution is not excluded.

@@

On the technical side, the discussion regarding the deployed experiment indicates directions for future investigation. The definition of a potential application will foster research aiming at passive or very low-energy devices that are able to effectively advertise their presence and collect statistics.

>>>>

This document addresses the specific aspect of heritage trail signage, and describes a solution based on QR code tags. The design gathers suggestions from several disciplines to take into account the many aspects that contribute to sustainability. The result is a practical signage technology that is adaptable to other environments sharing defined traits.

@@

The functions of signage are two:
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map. The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream". A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}). The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
 \item a sign should explain the reasons for interest in a site. This generic characterization includes all sorts of semantic characterization of the site, and may have educational or motivational intents. This includes describing historical facts related with the site, as well as advertising nearby points of interest (see figure \ref{fig:pole-direction}). In this sense, the CAI signal above is not sufficient. A simple board with a site name fits only when the site is sufficiently popular. Otherwise, a more structured message should be used to pursue an educational intent.
\end{itemize}

@@

This section outlines the historical framework that explains the selection of QR tag locations and their associated content. It is based on previous archaeological and historical investigations. The sites mentioned in the narration can be found among the QR codes listed in table \ref{tab:qrtags}.

@@

In Italy, a municipality is the smallest geographical-administrative unit offering accessible data about tourism, which are the fundamentals of our research.

%The municipality of \emph{Bagni di Lucca} boasts historical popularity for its landscapes told by famous poets (such as Giosuè Carducci, Giovanni Pascoli, Eugenio Montale, George Gordon Byron, Percy Bysshe Shelley) and celebrities like Paolina Bonaparte. The typical Liberty style architecture characterizes the villas like the \textit{Real Casinò} featuring elegant gardens representing the power of the Republic of \emph{Lucca} in the early XIX century and the \emph{Belle Epoque}. Such historical prominence made the town a well-known destination abroad, thanks to the massive emigration and the presence of the Thermal Bath structure \footnote{Terme di Bagni di Lucca: \url{https://www.termebagnidilucca.it/}}, nearby must-see attractions of the time.

@@

\begin{table}
    \footnotesize
    \setstretch{1,2}
    \centering
    \begin{tabular}{l||l|l} \hline 
        {\bf Region} & {\bf Bagni di Lucca} & {\bf Lucca} \\ \hline
        Resident density (residents/{$km^2$}) & 33.95 & 215.71 \\
        Touristic density (visitors/{$km^2$}) & 106.41 & 1866.96 \\
        Average stay  (days) &  2.98 &  3.36 \\
        Accommodation density (accommodations/{$km^2$}) & 10.14 & 19.30 \\
        Coexistence Index (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
         \hline
    \end{tabular}
    \caption{Tourist indexes in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
    \label{tab:tourism}
\end{table}

@@

Several promotional websites advertise Bagni di Lucca attractions such as these two destination websites: www.turismobagnidilucca.com and www.valdilima. org; the sitography closing this paper contains a long yet incomplete list that confirms such a statement.

From the point of view of our research, the context of Bagni di Lucca helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services). Indeed, Val di Lima can promote a very identifying destination brand straddling the town of Lucca and the Garfagnana mountain area. The Municipality of Bagni di Lucca and the tourist area of the Val di Lima are the context for the small and fascinating village of Casoli, which is one of the 31 villages belonging to the Municipality. It is 7.22 km far from Bagni di Lucca, at an altitude of 500 meters.

Casoli village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon park. Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities. Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices. After a detailed analysis of the tourist and social context of Bagni di Lucca, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{x} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experiences (\citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}).

As part of Bagni di Lucca's ecosystem, with its focus on tourism, Casoli is a destination for slow travel. It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric. This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{x}.

%Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade. Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley. Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long yet incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area. The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.
%
%From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).
%
%After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.
%
%Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area. 
%
%%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
%
%\begin{table}
%    \centering
%    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
%         {\bf Touristic promotion }&  {\bf Kind of service}\\
%         \hline \hline 
%         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
%         \hline 
%         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
%         \hline 
%         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
%         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
%        \hline 
%         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
%         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
%         \hline 
%         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
%         \hline 
%         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
%         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
%         \hline 
%        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
%        \hline
%    \end{tabular}
%    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players. The companies and associations in the list have been individually contacted but some did not respond to our survey on touristic information.}}}
%    \label{tab:experience}
%\end{table}
%
%The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality. It is 7,22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters of altitude.
%
%\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}. Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts. Definitely, despite its variety of tourist attractions, \emph{Casoli} exhibits unique features in its comprehensive tourism.
%
%%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
%
%Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways. According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities. Local stakeholders operating in the tourism system offer a variety of services all year round. Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability. As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination. Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric. The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.

@@

This section defines the points of interest selected to position the QR tags, and reports on their accessibility and state of conservation. This may be relevant in a perspective to spot the segment of users that may take advantage of the tag.

The cave named {\em Buca La Piella} is the foremost site from an archaeological point of view. It is located in a fascinating environment outside conventional circuits, which is why it is less frequented and consequently less compromised. It is reachable leaving a marked path to follow the bed of a tributary of the \emph{Lima} stream. The trail is dotted with minor karst caves, and the {\em Buca La Piella} is reachable by climbing up on a steep slope in the wood on the left of the watercourse.

@@

    \item URL LoR: such information is useful only in a few locations since the area is not uniformly covered by broadband networks: for instance, during a recon to the {\em Piella} cave, none of the smartphones of the participants received sufficient broadband network signal to visit the linked page. However, the application records the URL so that the user can visit it when entering a covered area. Each Web page contains site-specific information, an interactive map hosted by the uMap web service (\url{https://umap.openstreetmap.fr/it/}) with the location and content of all the QR tags (see the left box of figure \ref{fig:editing}), and a form for user feedback.
\end{itemize}

@@

Collecting statistics about user perception is not a primary interest for us at this stage. However, we provided two ways to collect such data. One is a 5-questions poll on the web pages linked to each tag. The user is asked to agreement a level of agreement with the following sentences:
\begin {itemize}
\item the QR tag was useful
\item a plaque would be preferable
\item I scanned other QR codes in the vicinity
\item I will come back to visit the sites I did not visit today
\item I am satisfied about my visit
\end{itemize}
In addition, the hits on the pages linked to the QR-codes were counted. Both ways proved to be ineffective. In the case of the form, we collected one positive feedback that has no statistical significance. The hit count was polluted by spurious access to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
%Its {\em performance} is currently difficult to evaluate. The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons, one being that the QR tag cannot record or transmit read events. Another is that the user may be unable or unwilling to visit the URL in the tag. Finally, the URL can be visited by means different from the QR reading. For such reasons, despite we acknowledging its importance for the stakeholders, we do not discuss such an index.

@@

Geo-itinerary

@@

A different application-based solution privileging passive devices is based on a writable NFC tag close to the one delivering the content. To update the number of visitors that have read the informational NFC, the user application reads and increments the count recorded on the writable NFC. Here too the primary drawback compared with our QR-tag based solution is the need for an ad-hoc application. In addition, the event of data pollution is not excluded.

@@

On the technical side, the discussion regarding the deployed experiment indicates directions for future investigation. The definition of a potential application will foster research aiming at passive or very low-energy devices that can effectively advertise their presence and collect statistics.
